---
title: "Fuel Consumption Analysis - Automatic vs. Manual Transmission"
output: 
author: "Polina Sklyarevsky"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

require(datasets)
require(GGally)
require(ggplot2)
require(car)
require(DescTools)
```

# Excecutive Summary

This report is based on the *mtcars* dataset, which includes various properties of 32 cars. The report shows that the transmission type (ehwther it's automatic or manual), the car's weight and the time it takes for a car to pass a quarter of a mile - are all statistically significant factors which the car's fuel consumption rate depends on (measured in miles per gallon). Using linear regression, this report concludes with 95% certainity that holding the abovementioned factors constant, **a car with manual transmission performs better in terms of fuel efficiency than a car with automatic transmission, "saving" about 3 miles per gallon**.

# The Data

```{r load_data}
data(mtcars)
```

```{r data_processing}
mtcars$am <- factor(x = mtcars$am, labels = c("Automatic", "Manual"))
mtcars$vs <- factor(x = mtcars$vs, labels = c("V", "S"))
```

The dataset consists of 32 cars, out of which 13 have Automatic Transmission ans 19 have Manual Transmission. 
- 'mpg' stands for "miles per gallon", and measured in miles (numeric).
- 'am' stands for Transmission Type, and measured in levels: automaric = 0; manual = 1 (factor). See appendix for full dataset description.

## Is an automatic or manual transmission better for MPG?

An indication of manual transmission efficiency, compared to automatic transmission, can be observed by looking at the mean values of mpg per transmission:

```{r summary_statistics}
summary(mtcars[mtcars$am == "Automatic",]$mpg)
summary(mtcars[mtcars$am == "Manual",]$mpg)
```

Apart from Transmission Type some other properties of the cars are correlated with mpg (see appendix for pairs plot and correlation values). A regression model will be used to further establish such correlations and test it's statistical significance. 

## Model Selection

The simplest *MPG ~ Transmission Type* model might require some adjustments, in prder to control for other variables which might also have a statisticaly significant effect on MPG. Therefore, a stepwise model selection process was performed (see appendix for full description). Both *weight* and *1/4 mile time* were found to have a statisticaly significant effect on MPG. Therefore, these variables were added to the following regression model.

## Results

**effect of Transmission on mpg, controlling for the weight of the car, and 1/4 mile time**

```{r results_wt+qsec}
summary(lm(formula = mpg ~ am + wt + qsec, data = mtcars))
```

From this model, it can be concluded with 95% certainity that Manual transmission is more effective than Automatic transmission, controlling for both the weight of the car and speed of acceleration (1/4 mile time). This model explains 84% of the variance in fuel consumption (MPG).

See appendix for model diagnostics.

## Counclusion

This report shows that controlling for both weight and 1/4 mile time of the car, it can be concluses with 95% certainty that **cars with Manual Transmission perform better in terms of fuel counsumpltion (about 3 MPG more) than cars with Automatic Transmission**.


## Appendix

### Dataset Description

```{r describe_dataset}
str(mtcars)
```

### Pairs Plot

```{r pairs_plot, message = FALSE}
ggpairs(data = mtcars, lower = list(continuous = "smooth"))
```

### Simple Regression Model

```{r basic_model}
m1 <- lm(formula = mpg ~ am, data = mtcars)
summary(m1)
```

### Model Selection 

```{r step_function}
m <- step(lm(formula = mpg ~ ., data = mtcars), trace = 0, steps = 1000)
summary(m)
```

```{r model_selection}
m2 <- update(object = m1, formula. = mpg ~ am + wt)
m3 <- update(object = m2, formula. = mpg ~ am + wt + qsec)

anova(m1, m2, m3)
```

It can be seen that when weight (wt) is added to the basic model, the new model performs significantly better. The same happens when adding 1/4 mile time (qsec) to the previous model.

Adding more variables to the basic model can inflate it's variance:
```{r variance_inflation_factors}
vif(m2)
vif(m3)
```

As can be seen, adding 1/4 mile time (qsec) to a model MPG ~ am + wt (our basic model + weight), causes variance inflation, but as VIF < 10 we'll consider it reasonable multicollinearity and move on with the last model.


### Model Diagnostocs

Residual plots for model: mpg ~ am + wt + qsec

```{r residual_plots_m3}
par(mfrow = c(2, 2))
plot(m3)
```
No outstanding patters can be seen from the above plots, suggesting that significant data was not left out. 

The normality of the residuals may look somewhat questionable, but Jarque-Berra test suggests normality (H0 of normality not rejected).

```{r normality}
JarqueBeraTest(m3$residuals)
```

### Code
can be found here: https://github.com/polinas123/regression
